---
output: github_document
---
[![Build Status](https://travis-ci.org/eribul/coder.svg?branch=master)](https://travis-ci.org/eribul/coder)
[![AppVeyor Build Status](https://ci.appveyor.com/api/projects/status/github/eribul/coder?branch=master&svg=true)](https://ci.appveyor.com/project/eribul/coder)
[![codecov](https://codecov.io/gh/eribul/coder/branch/master/graph/badge.svg)](https://codecov.io/gh/eribul/coder)


<!-- README.md is generated from README.Rmd. Please edit that file -->

```{r, echo = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>",
  fig.path = "README-"
)
```

# Note!

This package is work in progress! The repository is currently only used for 
internal development! More information will be available soon!

# Installation

```{r, eval = FALSE}
# install.packages("devtools")
devtools::install_github("eribul/coder")
```


# About the package

The coder package is used to classify items from one dataset, using code data from a secondary source. It was first developed for medical comorbidity data based on hospital visits recorded in a national patient register. Medical conditions were registered using standardized codes (ICD-10) and individual codes were summarized by weighting all individual comorbidities for each patient (the Charlson and Elixhauser comorbidity indices). Only hospital visits recorded within a specified time frame, compared to individual reference dates from the primary data source, were recognized as relevant.

The large data sets, as well as the complexity of the classification schemes make those calculations time consuming. A naïve approach using bare code comparisons and for-loops in R, took approximately 16 hours to run on a laptop computer with 16 GB of RAM. We were then able to reformulate the coding schemes using regular expressions and we optimized our package using the data.table package. The classification time were then reduced to a number of seconds. We also compared the coder package to two packages on CRAN with similar scoop, icd and comorbidities.icd10. Our package was considerably faster  than those.

The package does not only include classification schemes for comorbidity data. It incorporates a general framework for any case where items can be classified using standardized codes. It might therefore be relevant for many tasks involving data management in official statistics using big data. 

