---
title: "coder"
author: "Erik Bulow"
date: "`r Sys.Date()`"
output: rmarkdown::html_vignette
vignette: >
  %\VignetteIndexEntry{coder}
  %\VignetteEngine{knitr::rmarkdown}
  %\VignetteEncoding{UTF-8}
---

```{r setup, include = FALSE}
knitr::opts_chunk$set(
  collapse = TRUE,
  comment = "#>"
)
```

# Trivial example of car ownership

Let's assume we have a list of people, as well as dates for their driving license:

```{r}

library(coder)

people <- 
  data.frame(
    name = c(
      "Gavin, Christopher",
      "Albrecht, Lauren",
      "Aziz, Emily",
      "Svenne Petsson"
    ),
    license = as.Date(c(
      "1949-05-23",
      "2016-02-04",
      "2014-12-18",
      "1985-05-04"
    ))
  )
knitr::kable(people)
```

We then have a car registry with car ownership for 1,000 cars (where `code_date` specifies when the car was bought):

```{r}
knitr::kable(head(ex_cars))
```

We are no car freaks and therefore not interested in individual brand names of each car.
We just want to know who owns a car made by a specific company.
We can specify a classification scheme of brands related to each company (a `classcodes` object).
Note that not all models are classified since some brands were produced by companies that do no longer exist.
The classification scheme is made by regular expressions since these are more effective to work with for clasification of large data sets.


```{r}
knitr::kable(ex_carbrands)
```


# Categorize 

We now want to know if anyone owns a car that is still produced.

We use the three objects from above:

* `people`: the people we are interested in,
* `ex_cars`: the centralised car registry and
* `ex_carbrands`: our classification of car makers. 

Here we go:

```{r}
x <- categorize(people, ex_cars, ex_carbrands, id = "name")
knitr::kable(x)
```

# Interpretation

We can see that:

* Emily Aziz is a happy owner of a Mazda. 
* Lauren Albrecht owns both a Mercedes and a Toyota.
* Christopher Gavin might in fact own a car, but none of the important ones identified by our classification scheme.
* Svenne Petsson is a Swede and does not exist in our American car register.

The `index` column is just the sum of identified cars owned by each person.

Note that each column identifies if a person owns **at least one** car from the relevant company.


# Cars owned before license

We can use the date columns to check if anyone owned a car already before they got their drivers license.
We then need to specify that the relevant date column for this was the one called `license`.
The `days` argument specifies that we are interested in cars bought before this date.

```{r}
x <- categorize(people, ex_cars, ex_carbrands, id = "name", date = "license", days = c(-Inf, -1))
knitr::kable(x)
```

Oh, indeed, one of Lauren's car was actually bought before he could drive it himself. 


# Car rating

We are now not only interested in the number of cars owned by each individual. 
We would also like to rate the cars based on the manufacturer. Let's assign some points to each company.
We then specifies that the index should be calculated based on the new column named "points".

```{r}
ex_carbrands <- 
  as.classcodes(
    data.frame(
      ex_carbrands, 
      points = c(10, 1, 2, 1, 1, 1, 1, 2, 2, 5)
    )
  )

knitr::kable(ex_carbrands)

x <- categorize(people, ex_cars, ex_carbrands, id = "name", ind = "points")
knitr::kable(x)
```

It seems like, even though Emily owns just half as many cars as Albrecht, she actually gets more than twice the points when ranking the brands.


# Concluding remark

This vignette illustrates a simple example where the data sets are small, and the codnig schemes are simple. 
If that is the case, it might be equally good to just join or merge the data sets manually. 
A more realistic example of when the package might come handy is presented in the vignette called "Comorbidity".